---
author: "Nico Salm"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE, fig.height = 3)
library(tidyverse)
library(kableExtra)
source("../../scripts/viridis.R")
source("../../scripts/ggprob.R")
```

\newcommand{\E}{\mathsf{E}}
\newcommand{\Var}{\mathsf{Var}}
\renewcommand{\prob}{\mathsf{P}}


## Assignment 7

#### Due Saturday, October 28, 11:59 PM CT

### Preliminaries

- Directories
    - COURSE/homework/
    - COURSE/homework/hw07/
    - COURSE/data/
    - COURSE/scripts/
- Files
  - COURSE/homework/hw07/hw07.Rmd
  - COURSE/scripts/viridis.R
  - COURSE/scripts/ggprob.R

### Aims

- Practice probability distributions

### Problems

######  1.
A discrete random variable $X$ has possible values 0, 1, 2, 3, 4, 5, 6, 7, 8 with the following partial probability distribution.
The missing probabilities $\prob(X=7)$ and $\prob(X=8)$ are equal to each other.

```{r, echo = FALSE}
set.seed(20221021)
x = 0:8
p = rnorm(7,5,2.1)
p = sort(round(p / sum(p) * 0.84, 2))
p1 = tibble(x = x, p = c(p, rep(NA,2)))

p1_tab = p1 %>% 
  mutate(x = str_c(x)) %>% 
  pivot_wider(names_from = x, values_from = p) %>% 
  mutate(x = "P(X=x)", .before = `0`) 

p1_tab %>% 
  kable() %>% 
  kable_styling(position = "center", full_width = FALSE)
```

- Put the values `x = 0:8` and the tabled probabilities `p` into a data frame (use `tibble()`).

```{r}
p1 = tibble(x = x,
            p = c(p, rep(NA, 2)))

total_prob = sum(p1$p, na.rm = TRUE)
remaining_prob = 1 - total_prob
missing_prob = remaining_prob / 2

p1 = p1 %>%
  replace_na(list(p = missing_prob)) # this indeed sums to 1
p1
```

- Write code to calculate the missing values for $\prob(X=7) = \prob(X = 8)$ and replace the missing values with these calculated values into your data frame.

```{r}
sum_known_p = sum(p1$p, na.rm = TRUE)
remaining_p = 1 - sum_known_p
each_missing_p = remaining_p / 2
p1 = p1 %>%
  mutate(p = if_else(is.na(p), each_missing_p, p))
p1

```

- Verify that the sum of all probabilities is 1.

```{r}
sum(p1$p) # checks out
```



###### 2.
For the probability distribution from the previous problem:
  
- Calculate the mean, variance, and standard deviation.

```{r}
mean_x = sum(p1$x * p1$p)
var_x = sum((p1$x - mean_x)^2 * p1$p)
sd_x = sqrt(var_x)

c(mean_x, var_x, sd_x)
```




###### 3.
For the probability distribution from Problem 1:
  
- Draw a graph of the probability distribution with line segments at each possible value.
- Add to the graph a dashed red line at the mean and dotted black lines that are one and two standard deviations above and below the mean.
  
  
```{r}
# create basic scatter plot with line segments.
p = ggplot(data = p1, aes(x = x, y = p)) + 
  geom_point() +
  geom_segment(aes(x = x, xend = x, y = 0, yend = p)) # adds line segments

# add dashed red line for the mean.
p = p + geom_hline(yintercept = mean_x, linetype = "dashed", color = "red", size = 1)

# calculate positions for the standard deviation lines.
sd1_up <- mean_x + sd_x
sd1_down <- mean_x - sd_x
sd2_up <- mean_x + (2 * sd_x)
sd2_down <- mean_x - (2 * sd_x)

# add dotted lines for standard deviations.
p = p + geom_hline(yintercept = sd1_up, linetype = "dotted", color = "black", size = 1) +
      geom_hline(yintercept = sd1_down, linetype = "dotted", color = "black", size = 1) +
      geom_hline(yintercept = sd2_up, linetype = "dotted", color = "black", size = 1) +
      geom_hline(yintercept = sd2_down, linetype = "dotted", color = "black", size = 1)

p = p + labs(title = "Probability Distribution of X",
              x = "Value of X",
              y = "Probability")

p
```
  



###### 4.
For the probability distribution from Problem 1:
  
- Add to the `tibble` for the probability distribution for the random variable $X$ a column named `cdf` which contains the value $F(x) = \prob(X \le x)$.
    - Note that for this example, $F(0) = \prob(X = 0) = 0.04$ and $F(1) = \prob(X = 0) + \prob(X = 1) = 0.10$.
- Print the entire modified data frame.

```{r}
p1_with_cdf = p1 %>%
  arrange(x) %>% # ensure the data is sorted by 'x'.
  mutate(cdf = cumsum(p))

p1_with_cdf
```


The function $F(x)$ is known as the *cumulative distribution function*. In this example, $F(x) = 0$ for all $x < 0$, $F(x) = 1$ for all $x > 8$, and $0 < F(x) < 1$ if $0 < x < 8$. The function `cumsum()` may be very useful when calculating $F(x)$.
  


###### 5.
Using the probability distribution from the previous problem:
   
- Draw a graph to display the cumulative distribution function $F(x)$ with a step function.
    - For the purposes of the graph only, it may be helpful to add rows with $x=-1$ and $x=9$.
    - The **ggplot2** function `geom_step()` will be very useful to make this plot. See the help page or search the internet for an example on how to use it.
    
    
```{r}
extended_data <- p1_with_cdf %>%
  add_row(x = -1, cdf = 0) %>%  # Adding the starting point
  add_row(x = 9, cdf = 1) %>%  # Assuming the CDF is complete and reaches 1 at x = 9
  arrange(x)

ggplot(extended_data, aes(x = x, y = cdf)) +
  geom_step(direction = 'hv') + # 'hv' makes the step first horizontal then vertical
  labs(
    title = "Cumulative Distribution Function (CDF)",
    x = "x",
    y = "F(x)"
  )
```



###### 6.
The $p$ quantile of a probability distribution is any location $a$ where $\prob(X \le a) \ge p$ and $\prob(X \ge a) \ge 1-p$. For a continuous random variable with a density function, ther is typically a one-to-one correspondence between the quantile $a$ and the probability $p$. In contrast, for a discrete distribution such as the one featured in this assignment, the same value $a$ may be the quantile for many different $p$ and a few key values of $p$ have multiple possible quantile values $a$. The quantile function is essentially the inverse function $F^{-1}(p)$, which maps using the graph of $F$ from $p$ on the y-axis to the quantile $a$ on the x-axis.
  
- By examining the graph of the CDF in the previous problem, identify the 0.2, 0.3, 0.4, and 0.5 quantiles of the probability distribution of $X$. (You can do this without code, but a pat on the back if you can write code to find the values.)

```{r}
data <- tibble(X = runif(1000, 0, 8)) # generating random sample data

# Calculate quantiles
quantiles <- data %>%
  summarize(
    q0.2 = quantile(X, 0.2),
    q0.3 = quantile(X, 0.3),
    q0.4 = quantile(X, 0.4),
    q0.5 = quantile(X, 0.5)
  )

print(quantiles)
```

> 0.2 -> 1.63
> 0.3 -> 2.43
> 0.4 -> 3.30
> 0.5 -> 4.07
> (Calculated over a random sample)

- Explain why there are multiple values $a$ which are 0.1 quantiles of the distribution and specify all such values.

> This occurs when the CDF has a flat plateau. Therefore, all x-values in the range [0.0, slightly below 2.5] are 0.1 quantiles of the distribution.




###### 7.
The code below draws a graph of a discrete distribution called the *binomial distribution* with parameters $n = 5$ and $p = 0.4$. The median of the distribution is the 0.5 quantile. 
  
```{r}
gbinom(5, 0.4)
```
  
- Based on the graph, what value of $x$ does the median have? Compare to the value of `qbinom(0.5, 5, 0.4)`.

> The median is 2 which is also the output of qbinom(0.5, 5, 0.4)

```{r}
qbinom(0.5, 5, 0.4)
```



###### 8.
The code below draws a graph of a continuous normal distribution with parameters $\mu = 50$ and $\sigma = 10$ with the area to the left of 60 shaded.
  
```{r}
gnorm(50, 10) +
  geom_norm_fill(mu = 50, sigma = 10, b = 60) +
  scale_x_continuous(breaks = seq(30, 70, 10))
```
  
The value of `pnorm(60, 50, 10)` is the area to the left of 60 under this density curve.

```{r}
pnorm(60, 50, 10)
```

- Is the 0.80 quantile of this distribution smaller than, equal to, or greater than 60? Briefly explain.

> Since pnorm = 0.84 > 0.8, it means more than 80% of the data falls to the left of 60, indicating the the 0.80 quartile is less than 60.
    
 